Tuning Low Frequency Pulsar Searching with an Artificial Pulsar Device

### 2020-12-10 ms.tex
new file:   dm0_fb.pdf 	modified:   dm0_iq.pdf 	deleted:    dm0_profile.pdf 	modified:   dm1_iq.pdf 	deleted:    dm1_profile.pdf 	modified:   ms.tex
Updates from Overleaf

 \begin{figure}
    \centering
    \includegraphics[width=0.9\columnwidth]{dm1_profile.pdf}
    \caption{Frequency dependent pulse profile with a $\rm DM=1$ pc cm$^{-3}$ dispersion effect applied. The parameter of the pulsar
    is same as the one used in Figure \ref{fig:dm0_profile}}
    \label{fig:dm1_profile}
 \end{figure}

\subsubsection{Generating the narrow band I/Q sampling data}

With the amplitude profile in hand, we go on to generate the narrow band in-phase and quadrature (I/Q hereafter) signals.
First we determine the sampling interval $dt$ of the final generated I/Q time series signal and the number of frequency channels to $N_{\rm ch}$ be used in follow up procession.
The number of frequency channels further determines the bandwidth of each frequey channel as
\begin{gather}
    d\nu=\frac{f_{\max}-f_{\min}}{N_{\rm ch}}.
\end{gather}
For the sampling interval $dt$, according to the Nyquist Theorem the upper limit of which has to be $\frac{1}{2BW}$. 
For we are using I/Q sampling here the upper limit of the sampling interval is enlarged to $1/BW$.
The sampling interval of the narrow band I/Q data for current step can be calculated as
\begin{gather}
 dT=\frac{dt}{2N_{\rm ch}}.
\end{gather}
The factor 2 in the denominator is oversampling factor that we choose.
Thus the number of time bins within each period is determined 
\begin{gather}
 N_{\rm t}=\left\lfloor\frac{\tau}{dT}\right\rfloor.
\end{gather}
Note that in order to achieve a high computing performance, we limit the actual period $\tau$ can only be an integer multiple of $dT$.

So that for each frequency bin and time bin in a single period, we can calculate its instant complex voltage as
\begin{gather}
    U(\nu_{j}, t_{k})=\frac{1}{2}f^{\rm d}(\nu_j, t_{k}/\tau)(x^{\rm re}(j,k)+x^{\rm im}(j,k)i),
\end{gather}
where $j=1..N_{\rm ch}$, $k=1..N_{t}$, $i^2=-1$, and each $x^{\rm re}$ and $x^{\rm re}$ follows a standard normal distribution $\mathcal{N}(0,1)$ independently.

\subsubsection{Generating baseband I/Q sampling data stream}

The narrow band I/Q data is then fed into a over-sampled polyphase filterbank (PFB) to synthesize a broadband base-band I/Q data stream.
We take the work of \citet{2020JAI.....950004M} as a reference of designing the PFB.
We set the number of channels to be $N_{\rm ch}=16384$ and the taps of each channel to be 12.
PFBs are a kind of standard digital signal processing steps, so we just skip the details of the PFB that we use in this work.

The PFB outputs the baseband I/Q sampling time series.
We plot a sample of the baseband I/Q sample data in Figure \ref{fig:iq_data}(a), the DM of which is $1$ pc cm$^{-3}$.
It is obvious that all apparent time domain feature has been smeared because of the dispersion, so that pure time domain analysis is hardly possible to detect the fake pulsar we simulated here.
As a comparison, we present the same plot of the I/Q data with DM=0 pc cm$^{-3}$ in Figure \ref{fig:iq_data}(b), in which pulse-like time domain structure can be recognized clearly.
The I/Q sampling data stream is then sent to an SDR transmitter so that actual radio frequency (RF) signal within desired frequency range is generated continuously.
The RF signal is fed into a power amplifier and broadcast through a proper antenna.

\begin{figure}
    \centering
    \includegraphics[width=0.95\columnwidth]{dm1_iq.pdf}\\
    \includegraphics[width=0.95\columnwidth]{dm0_iq.pdf}
    \caption{The real part of the baseband I/Q data as a function of time. Two periods are shown here. The period is set to be 20 ms, in (a) the DM is set to be 10 pc cm$^{-3}$ and in (b), the DM is set to be 0 pc cm$^{-3}$ for the purpose of comparison.}
    \label{fig:iq_data}
\end{figure}

\subsection{Emitting Generated Signal}

\section{An Example of Tuning}

>>>>

 We present two examples of the time domain I/Q sampling data in Figure \ref{fig:iq_data} with ${\rm DM}=0$ pc~cm$^{-3}$ and DM=1 pc~cm$^{-3}$, respectively.
 The frequency range is set to be $100\sim 200$ MHz.
 Apparently the wide bandwidth and the low central frequency make the time domain data rather sensitive to the dispersion measure.
 ${\rm DM}=1$ is a rather small value of the dispersion measure among known pulsars.
 When trying to detect pulsars with larger dispersion measures in low frequency radio band, difficulties will appear caused by the the dispersion of ISM more seriously compared with higher radio bands.

\begin{figure*}
    \centering
    \includegraphics[width=0.95\columnwidth]{dm0_iq.pdf}
    \includegraphics[width=0.95\columnwidth]{dm1_iq.pdf}
    \caption{The real part of the baseband I/Q data as a function of time. Two periods are shown here. The period is set to be 20 ms, in (a) the DM is set to be 10 pc cm$^{-3}$ and in (b), the DM is set to be 0 pc cm$^{-3}$ for the purpose of comparison.}
    \label{fig:iq_data}
\end{figure*}

We show the waterfall chart of above generated time domain I/Q data in Figure \ref{fig:fb} to inspect the frequency-dependent profile.
The waterfall chart is generated by feeding the time domain I/Q data stream into a polyphase filterbank.
Apparently for the ${\rm DM}=1$ pc~cm$^{-3}$ signal, lower frequency has larger delay, which agrees with properties of dispersion effect caused by ISM. 
This verifies the correctness of the signal we generated.

\begin{figure*}
    \centering
    \includegraphics[width=0.9\columnwidth]{dm0_fb.pdf}
    \includegraphics[width=0.9\columnwidth]{dm1_fb.pdf}
    \caption{Frequency dependent pulse profile with a $\rm DM=0$ pc cm$^{-3}$ dispersion effect applied}
    \label{fig:fb}
\end{figure*}

\subsection{Emitting Generated Signal}
The generated broadband I/Q data stream that has been dispersed is then fed into some radio frequency transmitter.
We use the software defined radio (SDR) transceiver \textit{USRP} X310\footnote{\url{https://kb.ettus.com/X300/X310}} to send the generated signal.
The maximum sample rate can reach 200 Msps, while we use a sample rate of 100 Msps corresponding to a bandwidth of $B=100$ MHz.
The corresponding frequency range is then $[\nu_c-B/2, \nu_c+B/2]$, where $\nu_c$ is the central frequency.
The SDR transceiver is connected to a high performance workstation through a 10Gb Ethernet fiber.
In this work we set the central frequency $\nu_c=150$ MHz, so that the frequency range is $100\sim 200$ MHz.

There are other choices of devices that can be used to transmit the generated signal, e.g.,  \textit{LimeSDR}\footnote{\url{https://limemicro.com/products/boards/limesdr}}, 
\textit{HackRF}\footnote{\url{https://greatscottgadgets.com/hackrf}}, and other general purpose digital-analog converter cards. 
The reason why choose \textit{USRP} X310 is its maximum instant bandwidth of 200 MHz, which enables us to emulate the conditions of most of current working low frequency radio telescopes.

The generated radio frequency signal is then amplified and broadcast through an antenna of the corresponding frequency range.
Before broadcasting and receive the signal through antennas, we use an oscilloscope to check the voltage signal at the output terminal of the SDR transceiver as are shown in \ref{fig:osc_trace}.

\begin{figure}
    \centering
    \includegraphics[width=0.45\columnwidth]{dm_0_trace.eps}
    \includegraphics[width=0.45\columnwidth]{dm_0.05_trace.eps}
    \caption{The voltage signal trace observed with an oscilloscope at the output terminal of the SDR transceiver. The left and the right panel show signals with ${\rm DM}=0$ pc~cm$^{-3}$ and ${\rm DM}=0.05$ pc~cm$^{-3}$, respectively. The periods of both signals are set to be $20$ ms. Note that the lower limits of the frequency range of the generated signals are 100 MHz.}
    \label{fig:osc_trace}
\end{figure}

\section{Two Examples of Tuning}
We set up two experiments to demonstrate the process of tuning LFPODs to perform the pulsar signal detection. 
One of the experiment is using the SDR transceiver as a toy LFPOD.
Another experiment performed with the low frequency interferometer 21CMA.
For both experiment we follow the steps as
\begin{enumerate}
    \item Place the emitter (i.e., the \textit{USRP} X310 transceiver) close to the LFPOD;
    \item Set the emitter and the LFPOD to proper mode and start the data acquisition;
    \item After enough data is acquired and stored to disk, perform a pulsar search procedure to check whether the artificial pulsar signal can be recognized;
    \item If necessary, adjust the emitting setup.
\end{enumerate}

\subsection{HackRF---A Toy LFPOD}
\subsubsection{Setup of the experiment}
\textit{HackRF} is a low cost SDR transceiver with an operation frequency of 1 MHz$\sim$ 6 GHz.
The maximum sample rate is 20 MHz.
We place an \textit{HackRF} 1 m away from the emitter antenna to receive the signal.
As the distance is close enough, we do not use any external power amplifier either for the emitter and the receiver. 
The internal amplifier gains for the emitter and the receiver are set to 0 dB and 10 dB, respectively.
The sample rate of the receiver is set to 10 Msps, corresponding a bandwidth of 10 MHz and the central frequency is set to 150 MHz, which leads to a frequency range of $145\sim 155$ MHz.
The reason why we do not set the sample rate to its maximum allowed value is to prevent from packet dropping.

\subsubsection{Cases that we test}
We test two sets of artificial pulsar signal parameters, the periods of both are set to be 18 ms, and the DM's are respectively 10 pc~cm$^{-3}$ and 10 pc~cm$^{-3}$.

\subsubsection{Connecting the emitter and the receiver directly with a co-axis cable}

\subsubsection{Sending and receiving the signal through antennas}
Then we perform the data acquisition with the signal sent and received through antennas.
